The five pillars of computational reproducibility

### 2023-05-16 docs/5pillars.Rmd
minor works

@@

* Helps resolve issues.
These centralised platforms also enable discussions of code issues such as bugs and feature requests.
In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
and allows team members to raise potential issues with a colleagues code (or one's own).
This also allows the scientific community to raise an issue with the software maintainers.
For example, this feature is used extensively during the peer review process for the Journal of
Open-Source Software @Editorial2018-kt.

@@

Containers are an attempt to resolve some of the downsides of virtual machines and package managers.
Container images can be thought of as similar to virtual machines, but are more lightweight as they
share parts of the host operating system (**Figure 4**).
In contrast to VMs, running containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker, makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
containerisation system for web development and research.
Due to its early adoption, Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images called DockerHub.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

![Figure 4. Architecture of containers (left) compared to virtual machines (right)](../img/docker-containerized-and-vm-transparent-bg.webp)

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger, and cannot be shared in the code repository, so these need to be archived
in a long term repository, such as Zenodo or FigShare.

@@

When these features are brought together they facilitate extreme reproducibility, ie: makes reproducing a
previous study relatively easy in terms of complexity, number of commands, and in time.
For example, we made an example transcriptome analysis workflow based on the five pillars approach and
it could could be installed and generate differential expression and pathway enrichment results in a
matter of minutes @Ziemann_undated-zx.

@@

Some researchers have suggested to take it even further, by adapting software development
principles of unit tests and continuous integration/continuous delivery.
This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
made to the project's code base @Beaulieu-Jones2017-vj.

If the workflow executes successfully and tests are passed, the repository can be decorated with a
"badge" to inform users of the lack of errors in the code.
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves.
Therefore continuous analysis might be a future pediment (crowning feature) to the five-pillared
"temple" of reproducibl research.

@@

>>>>

Another challenge is the fact that different parts of the workflow have different compute needs.
For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
reference genome and variant calling which are demanding in memory and compute.
These computationally demanding steps are typically conducted on high end servers and workstations.
After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
so downstream analysis could be conducted on commodity PCs or laptops.
In those situations,

Later stage

@@

* Helps resolve issues.
These centralised platforms also enable discussions of code issues such as bugs and feature requests.
In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
and allows team members to raise potential issues with a colleagues code (or one's own).
This also allows the scientific community to raise an issue with the software maintainers.
For example, this feature is used extensively during the peer review process for the *Journal of
Open-Source Software* @Editorial2018-kt.

@@

Containers are an attempt to resolve some of the downsides of virtual machines and package managers.
Container images can be thought of as similar to virtual machines, but are more lightweight as they
share parts of the host operating system.
In the example shown in **Figure 4**, the six containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker, makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
containerisation system for web development and research.
Due to its early adoption, Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images called DockerHub.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

![Figure 4. Architecture of containers (left) compared to virtual machines (right) @noauthor_2021-lt](../img/docker-containerized-and-vm-transparent-bg.webp)

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger, and cannot be shared in the code repository.
They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
A better option for bioinformatics workflows would be to use the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in anarticle to a suitable persitent long-term archive (discussed below).
The recipe used to build the Docker image, called the Dockerfile, is sufficiently small and can be added
to the project repository.

@@

This work is not designed to give strong recommendations on the exact combination of tools or resources
to use in computational research, but to set out key principles and highlight some useful solutions
that users can incorporate into their toolkit to improve the overall reproducibility of their work.

When these five considerations are addressed, they facilitate extreme reproducibility, ie: makes
reproduction relatively quick, and simple.
For example, we made a transcriptome analysis workflow based on the five pillars approach and
it could could be installed and generate differential expression and pathway enrichment results in a
matter of minutes with only a few commands @Ziemann_undated-zx.

@@

Some researchers have suggested to take it even further, by adapting software development
principles of unit tests and continuous integration/continuous delivery.
This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
made to a project's code base @Beaulieu-Jones2017-vj.

If the workflow executes successfully and tests are passed, the repository can be decorated with a
"badge" to inform users of the lack of errors in the code.
If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
article can be assessed computationally, then such tests could be used to demonstrate computational
reproducibility without the need for individuals to execute the workflow themselves.
Therefore continuous analysis might be a future pediment (crowning feature) to the five-pillared
"temple" of reproducible research.

@@

There are other challenges worth mentioning.
Different parts of the workflow may have different compute needs.
For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
reference genome and variant calling which are demanding in memory and compute.
These computationally demanding steps are typically conducted on high end servers and workstations.
After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
so downstream analysis could be conducted on commodity PCs or laptops.
In those situations, a workflow manager such as Snakemake or Nextflow could be useful to
coordinate those hybrid computate workflows.
